---
title: "PubMed Counts of Articles"
author: "Joy Payton"
date: "2/18/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

What are the rates of publication of papers on various topics?

### R Markdown

This is an R Markdown File, which is a way to interlace three things:

* R (or other!) programming language
* Statistical or scientific reasoning about the code we write, and 
* The output of R code 

... all in one place.  You can export this interweaving of human and computer language as well as the code output to various formats -- pdf, Microsoft Word, or html.  Are you new to R Markdown?  Here are some great resources:

* R Markdown was developed by RStudio.  Read their descriptions and examples on [their site](https://rmarkdown.rstudio.com/).
* The RStudio gurus wrote a great book about it -- [check it out](https://bookdown.org/yihui/rmarkdown/)!

## Load Packages

Here we add the packages we'll use.  `tidyverse` helps reshape data, `easyPubMed` simplifies the use of the PubMed API, and `printr` allows us to show data frames more attractively. 

```{r packages}
library(tidyverse)
library(easyPubMed)
library(printr)
```

## A Single Search

I'd like to search across several years, for several different search terms.  Let's start simply, however, with a single search.  What can PubMed tell me about articles published in 2020 that mention "medicine" and "disparity"?

The results tell me the overall count (600) and give me the first few (from `RetStart` to `RetMax`) matching article IDs.

```{r single_example}
results <- get_pubmed_ids("2015[Date - Publication] AND medicine disparity")
results
```

I can also extract just the count of articles:

```{r show_count}
results$Count
```

But I don't want to do that for every combination of year and search term, to get the count of articles!

I'll start by creating my search terms.

## Create Search Terms

Let's create a data frame which contains all the combinations of the variables we want to search for.  First, we'll define three categories that we'll combine.  

* `years` will be the series from 2012 to 2022.  
* `terms` will be the terms we want to search on.

We'll make the combinatoric using `expand_grid()` from `tidyverse`.

```{r set_terms}
years <- c(2012:2023)

terms <- c(
  'medicine disparity', 
  'medicine racism', 
  'medicine racial bias')

search_terms <- expand_grid("year" = years,
                       "term" = terms)
```

Let's peek!

```{r look_terms_1}
head(search_terms, 10)
```

OK, now we'll pad those search terms with the text that the API requires:

```{r wrap_terms}
search_terms <- search_terms %>%
  mutate(final = paste(year, 
                       "[Date - Publication]",
                       " AND ",
                       term,
                       sep = ""
                       ))
```

And let's look again:

```{r look_terms_2}
head(search_terms, 20)
```

## Search in PubMed

Now we'll make a short function that returns the count of results for a given term:

```{r function}
count_results <- function(term) {
  results <- get_pubmed_ids(term)
  count <- as.integer(results$Count)
  return(count)
}
```

And now we'll use that function to populate a new column.  Note that we're using an `lapply` function that lets us put a pause between searches in order to not go over the "anonymous" API rate supported by PubMed.  

```{r populate_count}
search_terms <- search_terms %>% 
  mutate(num_results = lapply(final, function(f) {
    Sys.sleep(0.5)
    count_results(f)
    }))
```

## Visualize the Data

`ggplot` lets us take a look at our results graphically:

```{r visualize}
ggplot(search_terms, 
       aes(x=year, y=num_results)) +
  geom_col() +
  facet_wrap(term ~ .) +
  xlab("Year") +
  ylab("Count") + 
  ggtitle("Healthcare Disparity Articles")
```
